---
title: "Lecture 2: R basics"
subtitle: ""
author: "Sebastian Barfort"
date: "September, 2015"
output:
  ioslides_presentation:
    widescreen: true
    fig_caption: false
    highlight: zenburn
css: styles.css
---

## Today

- R basics
- groups
- exercise 1
 
## R basics

- do you all have R installed on your computer? 
- can you all read datasets into R?
- have you finished exercise 1?

## Installing packages

- on its own, R can't do all that much 
- to really make use of R's capabilities, we need packages
- a package bundles together code, data, documentation, and tests
- we install packages from two sources
    - the Comprehensive R Archive Network (CRAN)
    - github

## Installing packages from CRAN

We can install the `readr` package, for example, by running
    
```{r, eval = FALSE}    
install.packages("readr")
```

Afterwards, we can access all the functions available in the package by running

```{r, eval = FALSE}    
library("readr")
```

## Installing packages from Github

It's slightly more difficult to install from github since we need to load a package from CRAN first: `devtools`

Installing from Github now looks like

```{r, eval = FALSE}
library("devtools")
install_github("hadley/purrr")
```

the `purrr` package can now be loaded using the `library` command

```{r, eval = FALSE}
library("purrr")
```

## Getting help

- if you know the command, type `?` followed by the function in the console

```{r}    
?summary
```

- search your version of R using `??` followed by the function name
- otherwise, try Google/Stackoverflow or our discussion forum

## Fundamentals

R is *object oriented*

Examples: 

- character string (e.g. words)
- number
- vector
- matrix
- data frame
- list

We can verify the class of an object using the `class` function

We assign objects to object names using "`<-`" or "`=`"

## Example

```{r}
z = "text"
p = c(1, 3, 5)
q = 2
y = NA
k = FALSE
```

>- **Question**: what is the class of the objects given above?

## Special values

- `NA`: not avaliable, missing (`is.na`)
- `NULL`: undefined (`is.null`)
- `TRUE`: logical true (`isTRUE`)
- `FALSE`: logical false (`!isTRUE`)

## Symbols

| Operator | Meaning                  |
| -------- | ------------------------ |
| `<`      | less than                |
| `>`      | greater than             |
| `==`     | equal to                 |
| `<=`     | less than or equal to    |
| `>=`     | greater than or equal to |
| `!=`     | not equal to             |
| `a | b`  | a or b                   |
| `a & b`  | a and b                  |


## Functions

Functions operate on objects

R has many cool built in functions such as `summary`, `mean`, `table`, etc

```{r}
x = 1:10
mean(x)
sd(x)
table(x)
```

## Vectors

The most basic type of R object is a vector. There is really only one rule about vectors in R, which is that a vector can only contain objects of the same class.

Everything in R is a vector 

We can create vectors using the `c` (concatenate/combine) function 

```{r}
my_vector = c(1, 3, 5, 10)
another_vector = 1:100
a_third_vector = c("yes", "no", "hello")
my_logical_vector = c(TRUE, FALSE, FALSE, TRUE)
```

## Data frames

- R stores spreadsheet like data in a **data frame**
- these are really collections of vectors of the same length
- 



## survey results

```{r, message = FALSE, warning = FALSE}
#library("readr")
#
#df = read_csv("https://raw.githubusercontent.com/sebastianbarfort/sds/master/data/sds-su#rvey-1.csv")
```



# Data & Setup {.segue}


