---
layout: post
title:  "Topic 1: Reading And Working With Data"
categories: posts
---

In real life you will usually bring data into R from an outside file. This is rarely a pleasant experience, and we will devote several classes in a few weeks focused on how to clean and transform "messy" data. Since this is not our focus today, however, we will imagine all the data we work with today are "clean" (we will discuss what exactly "clean" and "messy" data is in the topic on data transformation). 

Usually, the data that you will work with can come from one of three sources[^datanote]

- statistics packages (data generated from Stata, SAS, SPSS, etc)
- Excel 
- flat files such as text files (`.txt`) or comma separated value files (`.csv`)

[^datanote]: data can also come from databases such as SQL, but this is too advanced for our course. 

Luckily, R has excellent capabilities for importing data from all sources listed above. I've created the following table to summarise the relevant options currently availible in R 

|Data|File Extension|Packages|Example|
|:----|:-----:|:-----|:-----|
|Statistical Packages| `.dta` (Stata) | `haven` | `read_dta("path/to/file")`
| | `.dta` (Stata) | `foreign` | `read.dta("path/to/file")`
| | `.sas7bdat` (SAS) | `haven` | `read_sas("path/to/file")`
| | `.sas7bdat` (SAS) | `sas7bdat` | `read.sas7bdat("path/to/file")`
|Excel| `.xlsx` | `readxl` | `read_excel("path/to/file")`
| | `.xlsx| `xlsx` | `read.xlsx("path/to/file")`
|Flat files| `.csv` | `base` | `read.csv("path/to/file")`
| |  | `readr` | `read_csv("path/to/file")`
| | `.txt` (SAS) | `base` | `read.table("path/to/file")`

The most common file format we will work with in this course is `.csv`. 

Let us try to read a `.csv` file into R. For normal sized data you will most likely store this in a folder on your computer. For this example, however, I will read a dataset on battles in the Game of Thrones series available online [here](https://github.com/chrisalbon/war_of_the_five_kings_dataset/blob/master/5kings_battles_v1.csv). 

I use the `readr` library to read the data. To access this library we run the `library("readr")` function. 


```{r}
library("readr")
```

Afterwards, we can read the data 
```{r}
got.df = read_csv("https://raw.githubusercontent.com/chrisalbon/war_of_the_five_kings_dataset/master/5kings_battles_v1.csv")
```

The `read_csv` function returns a data frame. There are many useful functions for getting a quick overview of the data. Here are just a few

- `names`: lists the column names of the data frame
- `head`: lists the first five observations 
- `summary`: calcultes summary statistics for the variables in the data frame
- `dim`: prints the number of rows and columns in the data frame

Let us start by examining the dimensions of the data

```{r}
dim(got.df)
```

By running the command we learned that the data frame contains xxx. 

Next, we print the column names

```{r}
names(got.df)
```

You can extract single variables and perform different operations on them. To extract a variable, we use the dollar sign (`$`). We can do more specific operations on this variable alone









