---
layout: post
title:  "Topic 1: Reading And Working With Data"
categories: posts
---

In real life you will usually bring data into R from an outside file. This is rarely a pleasant experience, and we will devote several classes in a few weeks focused on how to clean and transform "messy" data. Since this is not our focus today, however, we will imagine all the data we work with today are "clean" (we will discuss what exactly "clean" and "messy" data is in the topic on data transformation). 

What I want you to do before we meet is the following:

1. Open up RStudio and type along while reading the text below. Try to think about what we're doing instead of just zoning out and copying my code. 
2. Once you've done that, I want you to watch *Reading/Writing Data: Part 1 and 2* by Roger Peng available [here](https://www.youtube.com/watch?v=aBzAels6jPk&index=9&list=PLjTlxb-wKvXNSDfcKPFH2gzHGyjpeCZmJ).

Usually, the data that you will work with can come from one of three sources[^datanote]

- statistics packages (data generated from Stata, SAS, SPSS, etc)
- Excel 
- flat files such as text files (`.txt`) or comma separated value files (`.csv`)

[^datanote]: data can also come from databases such as SQL, but this is too advanced for our course. 

Luckily, R has excellent capabilities for importing data from all sources listed above. I've created the following table to summarise the relevant options currently available in R 

|Data|File Extension|Packages|Example|
|:----|-----:|-----|-----:|
|Statistical Packages| `.dta` (Stata) | `haven` | `read_dta("path/to/file")`
| | `.dta` (Stata) | `foreign` | `read.dta("path/to/file")`
| | `.sas7bdat` (SAS) | `haven` | `read_sas("path/to/file")`
| | `.sas7bdat` (SAS) | `sas7bdat` | `read.sas7bdat("path/to/file")`
|Excel| `.xlsx` | `readxl` | `read_excel("path/to/file")`
| | `.xlsx` | `xlsx` | `read.xlsx("path/to/file")`
|Flat files| `.csv` | `base` | `read.csv("path/to/file")`
| |  | `readr` | `read_csv("path/to/file")`
| | `.txt` | `base` | `read.table("path/to/file")`

<br> 

The most common file format we will work with in this course is `.csv`. 

Let us try to read a `.csv` file into R. For normal sized data you will most likely store this in a folder on your computer. For this example, however, I will read a dataset of battles in the Game of Thrones series available online [here](https://github.com/chrisalbon/war_of_the_five_kings_dataset/blob/master/5kings_battles_v1.csv). 

I use the `readr` library to read the data. To access this library we run the `library("readr")` function. The library must first be installed, however, which can be done by running the command `install.packages("readr")` or by accessing the Tools menu in RStudio, selecting Install Packages and typing in the name of the library.


```{r, warning = FALSE}
library("readr")
```

Afterwards, we can read the data 
```{r}
got.df = read_csv("https://raw.githubusercontent.com/chrisalbon/war_of_the_five_kings_dataset/master/5kings_battles_v1.csv")
```

The `read_csv` function returns a data frame. There are many useful functions for getting a quick overview of the data. Here are just a few

- `names`: lists the column names of the data frame
- `head`: lists the first five observations 
- `summary`: calcultes summary statistics for the variables in the data frame
- `dim`: prints the number of rows and columns in the data frame

Any R code following a hash (#) is not executed. These are called comments, and can and should be used to annotate and explain your code. For example, this doesn’t do anything.

```{r}
# this is a comment. R will not execute what I write here
```

Let us start by examining the dimensions of the data

```{r}
dim(got.df)
```

By running the command we learned that the data frame contains 38 rows and 25 columns. 

Next, we print the column names

```{r}
names(got.df)
```

You can extract single variables and perform different operations on them. To extract a variable, we use the dollar sign (`$`). We can do more specific operations on this variable alone.

The Game of Thrones data holds detailed information on every major battle. Let us create some summary statistics for the size of the attacker's force

```{r}
summary(got.df$attacker_size)
```

We note that there are 14 missing observations (`NA`s) and that the mean size of the attackers force is 9,943. 

The following build-in functions are useful for quickly summarizing a dataset

- `mean`
- `median`
- `min`
- `max`
- `quantile`

When we have non-numeric variables, the `table` function is useful for quick counts

```{r}
table(got.df$attacker_king)
```

We see that "Joffrey/Tommen Baratheon" is the most aggressive character in the dataset, accounting for 14 of the 38 attacks in the data. 

We can also get an idea of how succesfull each character is in their attacks by constructing a contingency table

```{r}
table(got.df$attacker_king, got.df$attacker_outcome)
```

We see that Stannis Baratheon was the least succesful character in the data as he lost as many battles as he won. 

Usually, it is much more productive to explore a dataset using visualisations instead of text and tables. In the next posts we will learn how to effectively visualise all kinds of data, but just to give you an idea of what is possible, I produce a few interesting plots of the data below. 

We will use the plotting package `ggplot2` for all the plotting in this course, so make sure this is installed if you want to run the code below

```{r, warning = FALSE, message = FALSE, fig.width=10, fig.height=6}
library("ggplot2")
p = ggplot(got.df, aes(x  = attacker_size, 
  fill = attacker_king))
p = p + geom_density(alpha = .6) # add density geom
p = p + scale_x_log10() # transform x axis to log scale
p = p + labs(x = "size of attacking army (log scale)", 
  title = "Whick king attacks with the largest army?")
plot(p)
```

```{r, warning = FALSE, message = FALSE, fig.width=10, fig.height=6}
p = ggplot(subset(got.df, attacker_king != ""), 
  aes( x = attacker_king))
p = p + geom_histogram() # add density geom
p = p + labs(x = "king", 
  title = "Which king was most active in the war?")
plot(p) 
```

## References

The Game of Thrones data used in this post is constructed by [Chris Albion](http://chrisalbon.com/). 
